S-Aligner: Ultra-scalable read mapping on Sunway Taihu Light

### 2017-04-03 response.txt

#############
To Reviewer1:
=============

Q1: The parallel mapping problem is trivial if the reference genome
can be replicated. This paper recognizes this, and partially
replicates the reference and the reads (this would be akin to
partitioning the machien into groups and running a set of reads on
each group with a replicated reference. This is a rather
straightforward approach.

A1: The top-level parallelism that uses replication is certainly
straightforward, but it was included in the paper for completeness.
All three levels of parallelism are critical for performance, albeit
some are more straightforward than others.

-------------

Q2: The other implementation techniques are relatively straightforward
adaptations of existing techniques (including the bitmapped matching
technique).

A2: While there certainly exist similar techniques, we believe the
novelty of the paper lies in the adaptation of these techniques for
the architecture and the application, which are unique.

#############

To Reviewer2:
=============

Q1: What are the main differences of the intra-CG parallelization of
S-Aligner with respect to the TPDS paper (see above)?

A1: While the Sunway architecture and the IBM Cell architecture share
similarities, looking at the detailed architectural characteristics
reveals a number of differences that are important to understand.  One
example is lack of a signaling system on the SW26010 processor which
makes a dynamic task scheduling system not possible (at least not very
easily).  Due to this we have to use a static task scheduling strategy
instead of the task queue implementation.  This leads to a number of
difficulties, which we were able to mitigate by dividing the kernel
into verification and alignment phases, and performing load balancing
between the two phases (Section 3.2 / Figure 8, at the bottom
alignment stage, the intervals are redistributed to each slave
processor, it does not cost much since the athread_spawn has minor
overhead).

-------------

Q2: Section 3.2/Figure 8: Where is the alignment result stored? How do
the SPs synchronize? The box at the bottom of the Slave Processor
Group is not explained.

A2: The alignment result is stored on the file system.  SPs
synchronize using athread_join, just like pthreads or CUDA threads.
The "Generate BP-wise alignment" phase happens only if an interval
passed the verification, while the intervals are redistributed to
other SPs after the athread_join of the verification phase.

-------------

Q3: Section 4.2/Table 1:

- Is the comparison between these two fair? The Sunway machine has
  more than twice the number of cores (albeit at less than half the
  frequency).

- When changing the size of reads, from 108 to 200 bps, RazerS3 takes
  only a fraction more, whereas the runtime for S-Aligner more than
  doubles. Why is that?

A3:

- We agree that the comparison is not direct apples-to-apples, but is
  included in the paper as a general guideline of what performance to
  expect from our work.  The x86 processors have a better performance
  at a give clock frequency (larger cache, better branch management,
  pipelining, and other processor front-end features), which makes our
  work even more appealing in comparison.

- RazerS3 does more optimization at the filtration stage while we
  focus on the optimization of verification stage.  The large cost
  difference at #bps = 200 is because we verify a much larger number
  of intervals.  We will clarify this in the paper.
  Also reads with length short than 200 is mostly used in short read
  mappers.

--------------
Q4: Section 4.3/Figure 14:

- Why start with 13 nodes?

- Do you consider the final phase of the application that gathers all
  results and writes them to a file?

A4:

- The reference is divided into 12 blocks (limitation of the size of
  the shared memory).  Each node processes one block while the
  remaining node loads reads.

- Yes, except that each process writes to a different file (instead of
  one common file, but each file is compatible to standard SAM/BAM viewer).
  So, there is no gathering of results and writing separate files is 
  more efficient on the Sunway file system.

---------------
To Review3:
===============

Q1: In Sect 4.1, the authors claim that making use of f SPs yielded a
speedup of 118X. That does not seem to match with the numbers in
Fig. 12. Could you explain how that speedup is computed?

A1: By dividing the computation of "MP Myers" and "SP Myers" (800 / 6.8 ~= 118).

---------------

Q2: In Table 1, when #bps is increased (in the last row), RazerS3
execution time is only increased by ~12% while there is a 2.9X
increase for S-Aligner. Does that suggest that S-Aligner is more
sensitive for increased bps values? What is the reason for that?  Can
this affect practical usability of the application (in real world use
cases)?

A2: Please refer to review2's A3.

---------------

Q3: Why scaling stopped at 13,312 nodes number of nodes?

A3: That was the access limit we had on the system.

===============

To The scratchpad size issue mentioned by Review 2:

athread_spawn has a very minor overhead in our implementation (less
than 1% of the computation time) so short task granularities are not a
major concern.  Further, our implementation does not require much
memory since it has O(n) space complexity and O(n^2) time complexity,
32KB of scrathpad memory is enough for a round of computation.

### 2017-04-03 s-aligner-ieee.tex
initial commit

\documentclass{sig-alternate-05-2015}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage[flushleft]{threeparttable}
\usepackage{multirow}
\usepackage{cases}

% REMOVE ME: added for revision
\usepackage{color}

% for the code listing
\usepackage{listings}
\lstset{language=C++,
    basicstyle=\ttfamily \footnotesize,
    keywordstyle=\color{black}\ttfamily,
    stringstyle=\color{black}\ttfamily,
    commentstyle=\color{gray}\ttfamily,
    morecomment=[s][\color{blue}]{/**}{**/},
    showstringspaces=false,
    showtabs=false,
    captionpos=b,
    numbers=left
}

@@

% Copyright
\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

% DOI
%\doi{10.475/123_4}

% ISBN
%\isbn{123-4567-24-567/08/06}

%Conference
%\conferenceinfo{PLDI '13}{June 16--19, 2013, Seattle, WA, USA}

%\acmPrice{\$15.00}

%
% --- Author Metadata here ---
%\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
% --- End of Author Metadata ---

\title{S-Aligner: Ultra-scalable read mapping on Sunway Taihu Light}
%\subtitle{[Extended Abstract]}
%\titlenote{A full version of this paper is available as
%\textit{Author's Guide to Preparing ACM SIG Proceedings Using
%\LaTeX$2_\epsilon$\ and BibTeX} at
%\texttt{www.acm.org/eaddress.htm}}}

\numberofauthors{7} 
%Xiaohui Duan, Kai Xu, Yuandong Chan, Christian Hundt,\\ Pavan Balaji, Bertil Schmidt, Weiguo Liu\\
\author{
\alignauthor 
Xiaohui Duan\\
    \affaddr{Shandong University}\\
    \affaddr{Jinan City, China}\\
    \email{sunrise.duan@mail.\\sdu.edu.cn}\\
\alignauthor
Kai Xu\titlenote{The first two authors contribute equally to the paper.}\\
    \affaddr{Shandong University}\\
    \affaddr{Jinan City, China}\\
    \email{xu\_kai@mail.sdu.edu.cn}\\
\alignauthor
Yuandong Chan\\
    \affaddr{Shandong University}\\
    \affaddr{Jinan City, China}\\
    \email{yuandong.c@mail.sdu.edu.cn}\\
\and
\alignauthor
Christian Hundt\\
    \affaddr{Johannes Gutenberg University}\\
    \affaddr{Mainz, Germany}\\
    \email{hundt@uni-mainz.de}\\
\alignauthor
Pavan Balaji\\
    \affaddr{Argonne National Laboratory}
    \affaddr{Argonne, Illinois, USA}\\
    \email{balaji@anl.gov}\\
\alignauthor
Bertil Schmidt\\
    \affaddr{Johannes Gutenberg University}\\
    \affaddr{Mainz, Germany}\\
    \email{bertil.schmidt@uni-mainz.de}\\
\and
\alignauthor
Weiguo Liu\\
    \affaddr{Shandong University}\\
    \affaddr{Jinan City, China}\\
    \email{weiguo.liu@sdu.edu.cn}\\
}

@@

\begin{abstract}
    The availability and amount of sequenced genomes have been rapidly growing in recent years due to the adoption of next-generation sequencing (NGS) technologies that allow for high-throughput short-read generation at highly competitive cost. Since this trend is expected to continue in the foreseeable future, the design and implementation of efficient and scalable NGS bioinformatics algorithms are important to research and industrial applications.  In this paper, we introduce S-Aligner---a highly scalable read mapper designed for the \emph{Sunway Taihu Light} supercomputer and its fourth-generation ShenWei many-core architecture (SW26010). S-Aligner employs a combination of optimization techniques to overcome both the memory-bound and the compute-bound bottlenecks in the read mapping algorithm. In order to make full use of the compute power of Sunway Taihu Light, our design employs three levels of parallelism: (1) inter-node parallelism using MPI based on a task-grid pattern, (2) intra-node parallelism using multi-threading and asynchronous data transfer to fully utilize all 260 cores of the SW26010 many-core processor, and (3) vectorization to exploit the available 256-bit SIMD vector registers.  Moreover, we have employed asynchronous access patterns and data sharing strategies during file I/O to overcome bandwidth limitations of the network file system. Our performance evaluation demonstrates that S-Aligner scales almost linearly with approximately 95\% efficiency for up to 13,312 nodes (concurrently harnessing more than 3 million compute cores). Furthermore, our implementation on a single node outperforms the established RazerS3 mapper running on a platform with eight Intel Xeon E7-8860v3 CPUs while achieving highly competitive alignment accuracy.

\end{abstract}

%
% The code below should be generated by the tool at
% http://dl.acm.org/ccs.cfm
% Please copy and paste the code instead of the example below. 
%
%\begin{CCSXML}
%<ccs2012>
% <concept>
%  <concept_id>10010520.10010553.10010562</concept_id>
%  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
%  <concept_significance>500</concept_significance>
% </concept>
% <concept>
%  <concept_id>10010520.10010575.10010755</concept_id>
%  <concept_desc>Computer systems organization~Redundancy</concept_desc>
%  <concept_significance>300</concept_significance>
% </concept>
% <concept>
%  <concept_id>10010520.10010553.10010554</concept_id>
%  <concept_desc>Computer systems organization~Robotics</concept_desc>
%  <concept_significance>100</concept_significance>
% </concept>
% <concept>
%  <concept_id>10003033.10003083.10003095</concept_id>
%  <concept_desc>Networks~Network reliability</concept_desc>
%  <concept_significance>100</concept_significance>
% </concept>
%</ccs2012>  
%\end{CCSXML}

%\ccsdesc[500]{Computer systems organization~Embedded systems}
%\ccsdesc[300]{Computer systems organization~Redundancy}
%\ccsdesc{Computer systems organization~Robotics}
%\ccsdesc[100]{Networks~Network reliability}

%
% End generated code
%

%
%  Use this command to print the description
%
%\printccsdesc

% We no longer use \terms command
%\terms{Theory}

\keywords{Sunway Taihu Light supercomputer; heterogeneous many-core architecture, bioinformatics, next-generation sequencing, read mapping}

@@

{\em Next-generation sequencing} (NGS) technologies have revolutionized the field of computational biology because of their massive throughput at highly competitive cost. Nowadays, NGS data is processed in almost every field of bioinformatics with important applications including personalized cancer treatment and precision medicine.
Modern NGS instruments such as Illumina sequencers \cite{reviewngs} allow for the recording of billions of short DNA fragments per day. The produced \emph{short reads} are usually only a few hundred base pairs (bps) in length compared with complete genomes of typical mammals covering billions of nucleotides. In order to further analyze the set of delocalized reads, they are typically mapped to a given reference genome by computing base-pair level alignments to determine their original positions. This exhaustive probing of possible alignments is a compute-heavy task and thus justifies the need for massively parallel and efficient computation patterns. Moreover, read alignment (or mapping) is crucial for many DNA sequence analysis pipelines such as genotyping\cite{genotyping}, the discovery of single nucleotide polymorphism\cite{snp}, or personal genomics\cite{personalgenomics}.

The na\"ive computation of optimal local alignments between each read and the reference genome is considered prohibitive because of the demanding time complexity of the Smith-Waterman algorithm \cite{sw} that is proportional to the product of the sequences' lengths. Thus, methods based on a {\em seed-and-extend} approach are typically employed to dramatically reduce the number of alignment positions by considering only those intervals in the reference genome which contain an exact occurrence of a substring of a read. Among existing read mappers one can find two predominant operation modes: {\em any-best} and {\em all}. Bowtie2 \cite{bowtie2}, BWA \cite{bwa}, GEM \cite{gem}, and CUSHAW \cite{cushaw} run in the {\em any-best} mode, which determines only one or a few best alignments (also called {\em intervals}) of each read to the reference genome. Other mappers such as RazerS3 \cite{razers3}, Hobbes2 \cite{hobbes2}, BitMapper \cite{bitmapper}, and mrFAST \cite{mrfast} operate in the {\em all} mode, which reports all intervals satisfying a certain constraint, for example, compliance with a predefined edit distance threshold.  Because of their extensive nature all-mappers demand far more computational resources than their any-best counterparts. Nevertheless, when studying gene innovation or phenotypic variation the complete list of alignments is crucial for a thorough analysis.

Consequently, we pursue the design of an all-mapper in order to support a wide range of biomedical applications. Associated runtimes can be dramatically reduced if we target clusters with a large number of compute nodes. In the recent past, the prevalent trend in supercomputing has favored heterogeneous nodes featuring attached coprocessors, such as GPUs or Xeon Phi architectures, because of their vast compute capabilities and low energy footprint. A similar architecture has recently been introduced with the \emph{Sunway Taihu Light} supercomputer, which is based on the neo-heterogeneous SW26010 many-core processor. A total of 10,649,600 cores from 40,960 nodes provide Linpack performance of up to 93 PFlop/s, making it the world's fastest supercomputer in the Top500 list of November 2016.

In this paper, we introduce S-Aligner---a highly scalable read mapper targeting the Sunway Taihu Light supercomputer and its SW26010 architecture. To the best of our knowledge, S-Aligner is the first attempt to implement a fully scalable read mapper specifically designed to fit the characteristics of Sunway Taihu Light. There are four major contributions in this paper. First, we provide an efficient implementation of Myers' bit-parallel string matching algorithm that harnesses the full parallelization and vectorization potential of the SW26010 many-core processor. Our new implementation is three orders of magnitude faster than a na\"ive single-threaded Myers' version. Second, we exploit fast local device memory via DMA intrinsics for intra-CG communication, which results in a 22$\times$ speedup compared with the non-DMA variant. Third, a highly scalable inter-CG communication scheme is proposed. This results in a parallelization efficiency of over 95\% when using 53,248 CGs. Fourth, we employ asynchronous file loading and data-sharing strategies to effectively hide the latency of the network file system. We further expect that the presented techniques can guide the design and implementation of similar types of applications on the neo-heterogeneous many-core cluster architecture of Sunway Taihu Light.

The rest of this paper is organized as follows: Section~\ref{Related Work} describes required background about the Sunway Taihu Light architecture, related work, and the seed-and-extend approach to read mapping. Details of the design and implementation of S-Aligner are presented in Section~\ref{Implementation}. Performance evaluations in terms of speed, scalability, alignment quality are carried out in Section~\ref{Evaluation}. Section~\ref{Conclusion} summarizes our conclusions.

@@

Each node is equipped with a single SW26010 processor that is subdivided into four {\em compute groups} (CGs). A CG is the basic unit that can be addressed by the job scheduling system. Each CG consists of a single {\em master processor} (MP), 64 {\em slave processors} (SPs), and 8 GB of attached DDR3 shared memory. All nodes can access a shared network file system for loading and storing data. The operating system is a customized Linux flavor running on the MP.  Users may manually launch threads on the SPs in order to parallelize compute-heavy portions of their code. Both the MP and 64 SPs support ShenWei's RISC basic instruction set, which provides scalar and SIMD operations. Moreover, compare and reduction primitives can be called exclusively on the MP, while the SPs exhibit specialized intrinsics for high-precision integer arithmetic. The 8 GB attached shared memory can be accessed by both the MP and the 64 SPs via a memory controller with a shared bandwidth of approximately 136 GB/s. The MP has its own cache (L1 and L2), and each SP can access 64 KB of fast {\em local device memory} (LDM). Data residing in shared memory can be written to LDM using DMA intrinsics and subsequently communicated via a broadcast to the LDM of other SPs. Figure~\ref{fig:SW26010} shows the described hardware layout.

@@

Previous algorithms mapped onto this novel supercomputer have focused mainly on application domains outside bioinformatics, such as Earth system modeling, ocean surface wave modeling, atomistic simulation, and phase-field simulation \cite{sunway}.

\subsection{Related Work}
There have been a lot of efforts to accerleate read mapping problem on clusters. Among the earliest tools are pBWA \cite{pbwa} and pMap \cite{pmap}. pBWA is an MPI implementation of Version 0.5.9 of the BWA aligner while pMap provides an MPI-based framework for the concurrent execution of popular aligners such as BWA \cite{bwa} and Bowtie \cite{bowtie}. Most of other implementations exploit MapReduce frameworks, for example, the Hadoop-based SEAL \cite{seal} and BigBWA \cite{bigbwa} tools or the Spark-based SparkBWA suite \cite{sparkBWA}, in order to guarantee fault tolerance. Unfortunately, all these approaches suffer from insufficient scalability when scaling to hundreds of nodes. The metaframework parSRA \cite{parSRA} is designed for read mappers written in UPC++.  It combines dynamic scheduling and a virtual file system layer for read distribution to overcome the limitations imposed by pMap and MapReduce-based approaches. merAligner \cite{merAligner} and CUSHAW3-UPC++ \cite{cushaw3upc} are other PGAS-based short-read aligners written in UPC++; both demonstrate good scalability on thousands of cores. Although pMap and parSRA allow for the execution of single node GPU aligners such as nvBowtie \cite{nvBio}, none of the cited tools is explicitly designed for heterogeneous clusters consisting of millions cores. Furthermore, all previous approaches target traditional hardware architectures.

\subsection{Seed-and-Extend approach}
%There are some "all" mappers currently on traditional commercial platforms, RazerS3\cite{razers3} is one of the most famous "all" mapper on x86\_64 CPUs. Hobbes2\cite{hobbes2}, BitMapper \cite{bitmapper} and mrFAST\cite{mrfast} are also read mappers in "all" mode on x86\_64 platforms. BitMapper2 also supports CUDA-enabled GPUs.

Consider a set of reads ${\cal R}$, a reference genome $G$, and an edit distance threshold $e$. The all-mode read mapping problem can be defined as follows: Find all substrings $g$ of $G$ whose edit distance to some read $R \in {\cal R}$ are no more than $e$. We call such occurrences $g$ in $G$ matches.

This problem can be solved by a classical {\em dynamic programming} (DP) approach that calculates the semi-global alignment between each $R \in {\cal R}$ and $G$. Unfortunately, each alignment results in a time complexity proportional to the product of the sequences' lengths, which renders intractable the alignment of a large number of short reads to a reference genome a few billion letters long.  In order to address this problem, most state-of-the-art solutions \cite{Reinert} employ the {\em seed-and-extend} approach, which follows a three-stage pipeline.

@@

\item[Stage 1: Filtration.] This stage identifies promising candidate intervals from $G$ (called {\em seeds}) for each read. A popular strategy in order to discard large intervals of $G$ is based on the fact that if a read $R \in {\cal R}$ is divided into $e+1$ non-overlapping $q$-grams (substrings of length $q = \lfloor \lvert R\rvert\ /(e+1) \rfloor$), then (according to the pigeonhole principle) at least one of them occurs exactly in a match. Such occurrences can be identified quickly by looking them up in precomputed $q$-gram index data structure (also called the {\em reference genome index}), which stores all substrings of length $q$ of $G$.

\item[Stage 2. Verification.] This stage determines whether a seed can actually be extended to a full match within edit distance threshold $e$. This requires the implementation of a {\em verification} algorithm in order to analyze the vicinity of each seed. Most mappers typically apply fast versions of DP-based algorithms for this step.

@@

Established all-mappers such as RazerS3 \cite{razers3} and mrFAST \cite{mrfast} also follow this pipeline. Our profiling of RazerS3 and mrFAST using a typical benchmark data set including the human reference HG19 and 1 million simulated Illumina reads (with a read length of 100 bps) reveals that Stage 2 occupies 67\% and 93\% of the overall runtime, respectively. This shows that a fast implementation of Myers' bit-parallel algorithm for computing the edit distance between a read and an interval is a crucial component when designing efficient all-mappers.

Consider two strings $s$ and $s \prime$ of length $n$ and $m$, respectively. Their {\em edit distance} is the minimum number of point mutations (i.e. insertions, deletions, or substitutions) required to transform $s$ into $s \prime$. It can be determined by relaxing the cells of a cost matrix $C$ of size $(n+1) \times (m+1)$ according to the recurrence relation shown in Equation~\ref{update_scheme}, where $0 < i \le n$ , $0 < j \le m$ and the {\em characteristic function} $\chi (x \ne y)$ is 1 if $x \ne y$, and 0 otherwise. Initial conditions are set to $C[i,0] = i$, $C[0,j] = j$, and $C[0,0] = 0$.

@@

Note that the absolute values in Equation~\ref{delta_update} are either 0 or 1. Thus, we can encode the relatively small state space with the help of five vectors using one-hot encoding. Figure~\ref{BitDP} shows an example for the encoding of the vertical derivative $\varDelta v$ and its associated one-hot representations $V^+$ and $V^-$.  With this bit-vector representation the cell updates can be rewritten in terms of logical operations as shown in Equation~\ref{one-hot-myers} \cite{myers}:

@@

Second, we evaluated the impact of using asynchronous filtration and the impact of using DMA intrinsics versus memcpy. The result is shown in Figure \ref{VarOptCha}.

>>>>

\documentclass[conference]{IEEEtran}

\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage[flushleft]{threeparttable}
\usepackage{multirow}
\usepackage{cases}
\usepackage{listings}
\lstset{language=C++,
    basicstyle=\ttfamily \footnotesize,
    keywordstyle=\color{black}\ttfamily,
    stringstyle=\color{black}\ttfamily,
    commentstyle=\color{gray}\ttfamily,
    morecomment=[s][\color{blue}]{/**}{**/},
    showstringspaces=false,
    showtabs=false,
    captionpos=b,
    numbers=left
}

\hyphenation{cluster computational-biology short-read-mapping}

@@

\title{S-Aligner: Ultra-scalable read mapping on Sunway Taihu Light}

\author{\IEEEauthorblockN{Xiaohui Duan, Kai Xu, Yuandong Chan}
\IEEEauthorblockA{%College of Computer Science and Technology \\
Shandong University\\
Jinan, China\\
Email: \{sunrise.duan, xu\_kai, yuandong.c\}@mail.sdu.edu.cn}
\and
\IEEEauthorblockN{Christian Hundt, Bertil Schmidt}
\IEEEauthorblockA{Johannes Gutenberg University\\
Mainz, Germany\\
Email: \{hundt, bertil.schmidt\}@uni-mainz.de}
\and
\IEEEauthorblockN{Pavan Balaji}
\IEEEauthorblockA{Argonne National Laboratory\\
Argonne, Illinois, USA\\
Email: balaji@anl.gov}
\and
\IEEEauthorblockN{Weiguo Liu}
\IEEEauthorblockA{Shandong University\\
Jinan, China\\
Email: weiguo.liu@sdu.edu.cn\\
}
}

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
The availability and amount of sequenced genomes have been rapidly growing in recent years due to the adoption of next-generation sequencing (NGS) technologies that allow for high-throughput short-read generation at highly competitive cost. Since this trend is expected to continue in the foreseeable future, the design and implementation of efficient and scalable NGS bioinformatics algorithms are important to research and industrial applications.  In this paper, we introduce S-Aligner---a highly scalable read mapper designed for the \emph{Sunway Taihu Light} supercomputer and its fourth-generation ShenWei many-core architecture (SW26010). S-Aligner employs a combination of optimization techniques to overcome both the memory-bound and the compute-bound bottlenecks in the read mapping algorithm. In order to make full use of the compute power of Sunway Taihu Light, our design employs three levels of parallelism: (1) inter-node parallelism using MPI based on a task-grid pattern, (2) intra-node parallelism using multi-threading and asynchronous data transfer to fully utilize all 260 cores of the SW26010 many-core processor, and (3) vectorization to exploit the available 256-bit SIMD vector registers.  Moreover, we have employed asynchronous access patterns and data sharing strategies during file I/O to overcome bandwidth limitations of the network file system. Our performance evaluation demonstrates that S-Aligner scales almost linearly with approximately 95\% efficiency for up to 13,312 nodes (concurrently harnessing more than 3 million compute cores). Furthermore, our implementation on a single node outperforms the established RazerS3 mapper running on a platform with eight Intel Xeon E7-8860v3 CPUs while achieving highly competitive alignment accuracy.
\end{abstract}

\IEEEpeerreviewmaketitle

@@

{\em Next-generation sequencing} (NGS) technologies have revolutionized the field of computational biology because of their massive throughput at highly competitive cost. Nowadays, NGS data is processed in almost every field of bioinformatics with important applications including personalized cancer treatment and precision medicine.
Modern NGS instruments such as Illumina sequencers \cite{reviewngs} allow for the recording of billions of short DNA fragments per day. The produced \emph{short reads} are usually only a few hundred base pairs (bps) in length compared with complete genomes of typical mammals covering billions of nucleotides. In order to further analyze the set of delocalized reads, they are typically mapped to a given reference genome by computing base-pair level alignments to determine their original positions. This exhaustive probing of possible alignments is a compute-heavy task and thus justifies the need for massively parallel and efficient computation patterns. Moreover, read alignment (or mapping) is crucial for many DNA sequence analysis pipelines such as genotyping, the discovery of single nucleotide polymorphism, or personal genomics.

The na\"ive computation of optimal local alignments between each read and the reference genome is considered prohibitive because of the demanding time complexity of the Smith-Waterman algorithm \cite{sw} that is proportional to the product of the sequences' lengths. Thus, methods based on a {\em seed-and-extend} approach are typically employed to dramatically reduce the number of alignment positions by considering only those regions in the reference genome which contain an exact occurrence of a substring of a read. Among existing read mappers one can find two predominant operation modes: {\em any-best} and {\em all}. Bowtie2 \cite{bowtie2}, BWA \cite{bwa}, GEM \cite{gem}, and CUSHAW \cite{cushaw} run in {\em any-best} mode, which determines only one or a few best alignments (also called {\em intervals}) of each read to the reference genome. Other mappers such as RazerS3 \cite{razers3}, Hobbes2 \cite{hobbes2}, BitMapper \cite{bitmapper}, and mrFAST \cite{mrfast} operate in {\em all} mode, which reports all intervals satisfying a certain constraint, for example, compliance with a predefined edit distance threshold.  Because of their extensive nature all-mappers demand far more computational resources than their any-best counterparts. Nevertheless, when studying gene innovation or phenotypic variation the complete list of alignments is crucial for a thorough analysis.

Consequently, we pursue the design of an all-mapper in order to support a wide range of biomedical applications. Associated runtimes can be dramatically reduced if we target clusters with a large number of compute nodes. In the recent past, the prevalent trend in supercomputing has favored heterogeneous nodes featuring attached coprocessors, such as GPUs or Xeon Phi architectures, because of their vast compute capabilities and low energy footprint. A similar architecture has recently been introduced with the \emph{Sunway Taihu Light} supercomputer, which is based on the neo-heterogeneous SW26010 many-core processor. This cluster consists of 40,960 nodes. Each node provides 260 compute cores organized in four {\em compute groups} (CGs). A total of 10,649,600 cores provides Linpack performance of up to 93 PFlop/s, rendering it the world's fastest supercomputer in the Top500 list of November 2016.

In this paper, we introduce S-Aligner---a highly scalable read mapper targeting the Sunway Taihu Light supercomputer and its SW26010 architecture. To the best of our knowledge, we are the first to investigate the applicability and scalability of read mapping on this architecture. Our approach relies on four major contributions. First, we provide an efficient implementation of Myers' bit-parallel string matching algorithm that harnesses the full parallelization and vectorization potential of the SW26010 many-core processor. Our new implementation is three orders of magnitude faster than a na\"ive single-threaded Myers' version. Second, we exploit fast local device memory via DMA intrinsics for intra-CG communication, which results in a 22$\times$ speedup compared with the non-DMA variant. Third, a highly scalable inter-CG communication scheme is proposed. This results in a parallelization efficiency of over 95\% when using 53,248 CGs. Fourth, we employ asynchronous file loading and data-sharing strategies to effectively hide the latency of the network file system. We further expect that the presented techniques can guide the design and implementation of similar types of applications on the neo-heterogeneous many-core cluster architecture of Sunway Taihu Light.

The rest of this paper is organized as follows: Section~\ref{Related Work} describes required background about the Sunway Taihu Light architecture, related work, and the seed-and-extend approach to read mapping. Details of the design and implementation of S-Aligner are presented in Section~\ref{Implementation}. Performance evaluations in terms of speed, scalability, and alignment quality are carried out in Section~\ref{Evaluation}. Section~\ref{Conclusion} summarizes our conclusions.

@@

Each node is equipped with a single SW26010 processor that is subdivided into four {\em compute groups} (CGs). A CG is the basic unit that can be addressed by the job scheduling system. Each CG consists of a single {\em master processor} (MP), 64 {\em slave processors} (SPs), and 8 GB of attached DDR3 shared memory. All nodes can access a shared network file system for loading and storing data. The operating system is a customized Linux flavor running on the MP.  Users may manually launch threads on the SPs in order to parallelize compute-heavy portions of their code. Both the MP and 64 SPs support ShenWei's RISC basic instruction set, which provides scalar and SIMD operations. Moreover, reduction primitives can be called exclusively on the MP, while the SPs exhibit specialized intrinsics for high-precision integer arithmetic. The 8 GB attached shared memory can be accessed by both the MP and the 64 SPs via a memory controller with a shared bandwidth of approximately 136 GB/s. The MP has its own cache (L1 and L2), and each SP can access 64 KB of fast {\em local device memory} (LDM). Data residing in shared memory can be written to LDM using DMA intrinsics and subsequently communicated via a broadcast to the LDM of other SPs. Figure~\ref{fig:SW26010} shows the described hardware layout.

@@

\subsection{Related Work}
We are not the first to parallelize read mapping on a compute cluster. Among the earliest tools are pBWA \cite{pbwa} and pMap \cite{pmap}. pBWA is an MPI implementation of Version 0.5.9 of the BWA aligner while pMap provides an MPI-based framework for the concurrent execution of popular aligners such as BWA \cite{bwa} and Bowtie \cite{bowtie}. Other implementations exploit MapReduce frameworks, for example, the Hadoop-based SEAL \cite{seal} and BigBWA \cite{bigbwa} tools or the Spark-based SparkBWA suite \cite{sparkBWA}, in order to guarantee fault tolerance. Unfortunately, all these approaches suffer from insufficient scalability when executed on hundreds of nodes. The metaframework parSRA \cite{parSRA} is designed for read mappers written in UPC++.  It combines dynamic scheduling and a virtual file system layer for read distribution to overcome the limitations imposed by pMap and MapReduce-based approaches. merAligner \cite{merAligner} and CUSHAW3-UPC++ \cite{cushaw3upc} are other PGAS-based short-read aligners written in UPC; both demonstrate good scalability. Although pMap and parSRA allow for the execution of single node GPU aligners such as nvBowtie \cite{nvBio}, none of the cited tools is explicitly designed for heterogeneous clusters consisting of thousands of many-core processors. Furthermore, all previous approaches target traditional hardware architectures.

To the best of our knowledge, S-Aligner is the first attempt to implement a fully scalable read mapper specifically designed to fit the characteristics of Sunway Taihu Light. Previous algorithms mapped onto this novel supercomputer have focused mainly on application domains outside bioinformatics, such as Earth system modeling, ocean surface wave modeling, atomistic simulation, and phase-field simulation \cite{sunway}.

\subsection{Seed-and-Extend Approach}
%There are some "all" mappers currently on traditional commercial platforms, RazerS3\cite{razers3} is one of the most famous "all" mapper on x86\_64 CPUs. Hobbes2\cite{hobbes2}, BitMapper \cite{bitmapper} and mrFAST\cite{mrfast} are also read mappers in "all" mode on x86\_64 platforms. BitMapper2 also supports CUDA-enabled GPUs.

Consider a set of reads ${\cal R}$, a reference genome $G$, and an error threshold $e$. The read mapping problem can be defined as follows: Find all substrings $g$ of $G$ that are within edit distance $e$ to some read $R \in {\cal R}$. We call such occurrences $g$ in $G$ {\em matches}.

This problem can be solved by a classical {\em dynamic programming} (DP) approach that calculates the semi-global alignment between each $R \in {\cal R}$ and $G$. Unfortunately, each alignment results in a time complexity proportional to the product of the sequences' lengths, which renders intractable the alignment of a large number of short reads to a reference genome a few billion letters long.  In order to address this problem, most state-of-the art solutions \cite{Reinert} employ the {\em seed-and-extend} approach, which follows a three-stage pipeline.

@@

\item[Stage 1: Filtration.] This stage identifies promising candidate regions (called {\em seeds}) for each read in $G$. A popular strategy in order to discard large regions of $G$ is based on the fact that if a read $R \in {\cal R}$ is divided into $e+1$ non-overlapping $q$-grams (substrings of length $q = \lfloor \lvert R\rvert\ /(e+1) \rfloor$), then (according to the pigeonhole principle) at least one of them occurs exactly in a match. Such occurrences can be identified quickly by looking them up in precomputed $q$-gram index data structure (also called the {\em reference genome index}), which stores all substrings of length $q$ of $G$.

\item[Stage 2. Verification.] This stage determines whether a seed can actually be extended to a full match within edit distance $e$. This requires the implementation of a {\em verification} algorithm in order to analyze the vicinity of each seed. Most mappers typically apply fast versions of DP-based algorithms for this step.

@@

Established all-mappers such as RazerS3 \cite{razers3} and mrFAST \cite{mrfast} follow this pipeline. Our profiling of RazerS3 and mrFAST using a typical benchmark data set including the human reference HG19 and 1 million simulated Illumina reads (with a read length of 100 bps) reveals that Stage 2 occupies 67\% and 93\% of the overall runtime, respectively. This shows that a fast implementation of Myers' bit-parallel algorithm for computing the edit distance between a read and an interval is a crucial component when designing efficient all-mappers.

Consider two strings $s$ and $s^\prime$ of length $n$ and $m$, respectively. Their {\em edit distance} is the minimum number of point mutations (i.e. insertions, deletions, or substitutions) required to transform $s$ into $s^\prime$. It can be determined by relaxing the cells of a cost matrix $C$ of size $(n+1) \times (m+1)$ according to the recurrence relation shown in Equation~\ref{update_scheme}, where $0 < i \le n$ , $0 < j \le m$ and the {\em characteristic function} $\chi (x \ne y)$ being 1 if $x \ne y$, and 0 otherwise. Initial conditions are set to $C[i,0] = i$, $C[0,j] = j$, and $C[0,0] = 0$.

@@

Note that the absolute values in Equation~\ref{delta_update} are either 0 or 1. Thus, we can encode the relatively small state space with the help of five vectors using one-hot encoding. Figure~\ref{BitDP} shows an example for the encoding of the vertical derivative $\varDelta v$ and its associated one-hot representations $V^+$ and $V^-$.  With this bit-vector representation the cell updates can be rewritten in terms of logical operations as shown in Equation~\ref{one-hot-myers}:

@@

Second, we evaluated the impact of using asynchronous filtration and the impact of using DMA intrinsics versus \texttt{memcpy}. The result is shown in Figure \ref{VarOptCha}.
